---
title: "newtidyingcleaning"
format: html
editor: visual
---
---
title: "Cleaning & Wrangling"
format: html
editor: visual
editor_options: 
  chunk_output_type: inline
---

```{r}

#| message:false
#| 
library(dplyr)
library(tidyr)
library(readr)

```

# Data Sources

## read in data

```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## explore data

```{r}
# print colum names
colnames(catch_original)
head(catch_original)

## Summary of each column of data
summary(catch_original)

## Prints unique values in a column (in this case, the region)
unique(catch_original$Region)


```
## cleaning and reshaping

##use the select()function
```{r}
catch_data <- catch_original %>%
    select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

head(catch_data)

catch_data <- catch_original %>%
    select(-All,-notesRegCode)

head(catch_data)
```
##check colum type

glimpse(catch_data)

##replace values using mutate
##change chinook 
```{r}
catch_clean <- catch_data %>%
    mutate(Chinook = as.numeric(Chinook))``{r}

```
```{r}
i <- which(is.na(catch_clean$Chinook))
i
```
```{r}
catch_clean <- catch_data %>%
    mutate(Chinook = as.numeric(Chinook))
i <- which(is.na(catch_clean$Chinook))
i

catch_clean <- catch_data %>%
    mutate(Chinook = if_else(condition = Chinook == "I", 
                             true = "1", 
                             false = Chinook),
           Chinook = as.numeric(Chinook))

##check
catch_clean[i, ]



```



```
## pivot the data shape using pivot_longer and pivot_wider

```{r}
catch_long <- catch_clean %>% 
    #pivot longer all columns except Region and Year
    pivot_longer(
        cols = -c(Region, Year),
        names_to = "species",
        values_to = "catch")

head(catch_long)

```
```{r}
catch_long <- catch_clean %>% 
    #pivot longer all columns except Region and Year
    pivot_longer(
        cols = -c(Region, Year),
        names_to = "species",
        values_to = "catch")

head(catch_long)
```
## pivot wide
```{r}

catch_wide <- catch_long %>%
    pivot_wider(names_from = species,
                values_from = catch)

head(catch_wide)
```

## renaming columns using rename()
```{r}

catch_long <- catch_long %>%
    rename(catch_thousands = catch)

head(catch_long)
```

```{r}
catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000)

head(catch_long)
```

```{r}
catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000) %>%
    select(-catch_thousands)

head(catch_long)
```
```{r}
unique(catch_original$Region)
```

```{r}
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch))

head(mean_region)
```

```{r}

n_region <- catch_long %>%
    group_by(Region) %>%
    summarize(n = n())

head(n_region)
```
```{r}

sse_catch <- catch_long %>%
    filter(Region == "SSE")

head(sse_catch)
```
```{r}
catch_long %>%
  filter(Region == c("SSE", "ALU")) %>%
  nrow()
```

```{r}
catch_long %>%
  filter(Region %in% c("SSE", "ALU")) %>%
  nrow()
```

```{r}

catch_million <- catch_long %>%
    filter(catch > 1000000)
```

chinook_sse <- catch_long %>%
    filter(Region == "SSE",
           species == "Chinook")

chinook_sse <- catch_long %>%
    filter(Region %in% "SSE" & species %in% "Chinook")
    head(chinook_sse)
```
## arrange by region
```{r}

mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch)) %>%
    arrange(mean_catch)

head(mean_region)
```

##sort by decending order
```{r}

mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch)) %>%
    arrange(desc(mean_catch))

head(mean_region)
```

##now separate and unite
```{r}
sites_df <- data.frame(site = c("HAW-101",
                                "HAW-103",
                                "OAH-320",
                                "OAH-219",
                                "MAU-039"))

sites_df %>%
    separate(site, c("island", "site_number"), "-")

```


## new cities
```{r}
## create `cities_df`
cities_df <- data.frame(city = c("Juneau AK",
                                 "Sitka AK",
                                 "Anchorage AK"))

```
## split into city and state

```{r}
colnames(cities_df)

cities_clean <- cities_df %>%
    separate(city, c("city", "state_code"), " ")
```
##unite funcction
```{r}

dates_df <- data.frame(
    year = c("1930",
             "1930",
             "1930"),
    month = c("12",
              "12",
              "12"),
    day = c("14",
            "15",
            "16")
)

dates_df %>%
    unite(date, year, month, day, sep = "-")
```
## now real analysis
```{r}

catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")

mean_region <- catch_original %>%
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = if_else(Chinook == "I", "1", Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year), 
               names_to = "species", 
               values_to = "catch") %>%
  mutate(catch = catch * 1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))

head(mean_region)
```

## now all together 
```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
mean_region <- catch_original %>%
select(-All, -notesRegCode) %>% 
  mutate(Chinook = if_else(Chinook == "I", "1", Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year), 
               names_to = "species", 
               values_to = "catch") %>%
  mutate(catch = catch * 1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))
head(mean_region)
```
## read out a csv
```{r}
write_csv(mean_region, here::here("data/mean_catch_by_region.csv"))

```

