---
title: "Cleaning & Wrangling"
format: html
editor: visual
editor_options: 
  chunk_output_type: inline
---

```{r}

#| message:false
#| 
library(dplyr)
library(tidyr)
library(readr)

```

# Data Sources

## read in data

```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## explore data

```{r}
# print colum names
colnames(catch_original)
head(catch_original)

## Summary of each column of data
summary(catch_original)

## Prints unique values in a column (in this case, the region)
unique(catch_original$Region)


```
## cleaning and reshaping

##use the select()function
```{r}
catch_data <- catch_original %>%
    select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

head(catch_data)

catch_data <- catch_original %>%
    select(-All,-notesRegCode)

head(catch_data)
```
##check colum type

glimpse(catch_data)

##replace values using mutate
##change chinook 
```{r}
catch_clean <- catch_data %>%
    mutate(Chinook = as.numeric(Chinook))``{r}

```
```{r}
i <- which(is.na(catch_clean$Chinook))
i
```
```{r}
catch_clean <- catch_data %>%
    mutate(Chinook = as.numeric(Chinook))
i <- which(is.na(catch_clean$Chinook))
i

catch_clean <- catch_data %>%
    mutate(Chinook = if_else(condition = Chinook == "I", 
                             true = "1", 
                             false = Chinook),
           Chinook = as.numeric(Chinook))

##check
catch_clean[i, ]



```



```
## pivot the data shape using pivot_longer and pivot_wider

```{r}
catch_long <- catch_clean %>% 
    #pivot longer all columns except Region and Year
    pivot_longer(
        cols = -c(Region, Year),
        names_to = "species",
        values_to = "catch")

head(catch_long)

```
```{r}
catch_long <- catch_clean %>% 
    #pivot longer all columns except Region and Year
    pivot_longer(
        cols = -c(Region, Year),
        names_to = "species",
        values_to = "catch")

head(catch_long)
```
## pivot wide
```{r}

catch_wide <- catch_long %>%
    pivot_wider(names_from = species,
                values_from = catch)

head(catch_wide)
```

## renaming columns using rename()
```{r}

catch_long <- catch_long %>%
    rename(catch_thousands = catch)

head(catch_long)
```
